Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-02-17 bbbq_article.tex
Process feedback @bogeert. Thanks Geert! Progress #179

It is helpful to be able to predict which peptides are immunogenic,
in, among others, vaccine development. 
Already for a decade, synthesized peptide fragments are used
to experimentally determine which peptides
are immunogenic.
This approach, however, is tedious and costly.
Due to this, software was written that allows for silico 
predictions, which are reliable in 
practice (\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).

@@

As the immunoproteasome is a cytosolic protein complex, on expects it
to cleave only cytosolic proteins. However, it has recently been
discovered that transmembrane protein residues are also presented
by MHC-I complexes (\cite{bianchi2017}). This discovery is based
on both in vitro elution studies, as well as on in silico predictions,
with a focus on transmembrane helices. 
The intracellular pathway, in which transmembrane residues are loaded onto
an MHC, however, is yet unknown.

Transmembrane helices (TMHs) 
are structurally conserved and hydrophobic 
regions of a protein that span a (also hydrophobic) 
cell membrane by an alpha helix of usually 23 amino acids.
Together with the internals of soluble proteins,
peptides derived from TMHs are the most hydrophobic
of a proteome. In both cases, the hydrophobic regions
help shape the protein in its correct 3D configuration.
TMHs are common structures in the proteins of
humans and pathogens. For example, 
15-39\% of the proteins in humans
is predicted to be a membrane protein (\cite{ahram2006estimation}).
There are multiple tools to do in silico predictions 
to determine the transmembrane regions of a protein 
(\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}).

@@

For MHC-I, it was found that predicted epitopes derived 
from human transmembrane helices (TMHs)
are over-presented by all 5 HLA-A and 
most of 8 HLA-B super-types (\cite{bianchi2017}),
due the high hydrophobicity as is required for a TMH to span
the (hydrophobic) cell membrane, combined with the fact that the MHC
binding cleft is hydrophobic as well.

It is unknown if the presentation of hydrophobic/TMH residues
is evolutionary selected for.
One reason to expect a reduced variability (and hence 
evolutionary conservation) in TMHs, is
that these are restricted in their evolution
by the functional requirement to span a lipid bilayer.
Due to this, the TMHs that pathogens possess, 
have a lower chance to develop an escape mutation,
as many mutations will result in a dysfunctional TMH.

% \paragraph{Does MHC-I present TMH-derived epitopes from pathogens as often?}

It is important that MHC-I presents both the peptides of the 
(healthy) cell, as well as possible pathogen-derived peptides when the
cell is infected. As described above, MHC-I presents TMH-derived epitopes 
from the human host more often. It is unknown if MHC-I has the same
dedication to present epitopes that stem from TMHs derived from
proteins produced by pathogen, either would the pathogen
be a virus or 
a bacterium.

% \paragraph{MHC-II is expected to present TMHs}

If presentation of TMHs on MHC-I would bring an evolutionary advantage 
in the recognition of pathogens by the immune system, 
it would follow that this is equally important for MHC-II, 
especially as the help of CD4+ T cells is needed for a long lasting CD8+ T cell 
response (\cite{novy2007cd4}). 
The mechanism to detect foreign TMHs by MHC-II would be unknown, 
similar to the discovery that MHC-I presents TMHs.

% \paragraph{Does MHC-II present TMH-derived epitopes from pathogens as often?}

It is unknown if MHC-II, like MHC-I, 
presents TMH-derived epitopes as often, either 
in humans,
bacteria or viruses.

% \paragraph{Selection undetectable in whole proteome}

In general, on would hope that evolutionary selection results in
an immune system that as most attentive for loci that are
essential for a virus, as these will be most conserved.
In SARS-CoV-2, for example, there is preliminary evidence that the strongest
selection pressure is upon residues that changes its 
virulence (\cite{velazquez2020positive}).
These loci, however, only account for a small part of a pathogen's proteome.
Additionally, these essential parts differ widely between pathogens.
Because of this scarcity and variance in targets, 
one can imagine that the human immune system 
is not tailored to detect these sites, 
as hinted by upon by the aforementioned influenza study.

% \paragraph{Selection may be detectable in TMHs}

TMHs, on the other hand, also have their function constraints, 
yet can occur multiple time a pathogen's proteome.
One can safely assume a pathogen's proteome contains multiple TMHs.
Therefore, it may be beneficial for the host
if its immune system would be more attentive towards TMHs.
And maybe this has already happened: MHC-I already detects hydrophobic
peptides. This feature, however, may also be caused by selection
to detect hydrophobic regions in the soluble proteins of pathogens.
It is unknown, when focusing on TMHs only, if a signal of selection
can be detected.

% \paragraph{Use of protein data in phylogenetics}

When using DNA sequences, one can use a skewed rate
in non- versus synonymous mutation
to detect the signature of selection (\cite{murrell2015gene}).
Using AA sequences, however, has its advantages,
as its is closer to the actual phenotype
selection acts upon: DNA may never be transcribed to RNA,
or its RNA may never be translated (\cite{diz2012proteomics}).
When using a proteome in phylogenetic research,
we know that the majority of proteins are selected to just
maintain their function most of the time, where
is the time spans there is selection, only a few AAs
can actually increase the 'fitness' of the
protein (\cite{anisimova2009investigating}).
There, when generalizing the dynamics of mostly purifying selection (to maintain
a protein's function) and a short duration of positive selection,
those genes that are selected cannot be detected (\cite{yang2000statistical}).

>>>>

Much studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development 
of very reliable prediction algorithms 
that allows for in silico predictions 
of the binding affinities of peptides (\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).

@@

Using these prediction algorithms, 
we recently showed that peptides derived 
from transmembrane helices (TMHs) 
are predicted to be more frequently presented by MHC-I 
than expected based on their abundance (\cite{bianchi2017}). 
Moreover, we showed that some well-known immunodominant peptides stem from TMHs.

This over-presentation is entirely explained by the fact 
that the peptide-binding groove of most MHC-I haplotypes 
is relatively hydrophobic, 
and therefore hydrophobic peptides have a higher affinity 
than hydrophilic peptides.

Transmembrane helices (TMHs) are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes, 
and they consist of an alpha helix of usually 23 amino acids. 
TMHs are common structures in the proteins of humans and pathogens. 
For example, 
15-39% 
\richel{Why wide range? Issue 180}
of the proteins in humans 
are predicted to be a membrane protein 
containing at least one TMH (\cite{ahram2006estimation}). 
TMHs can also be predicted from a protein sequence 
with high accuracy by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}).

@@

%For MHC-I, it was found that predicted epitopes derived 
%from human transmembrane helices (TMHs)
%are over-presented by all 5 HLA-A and 
%most of 8 HLA-B super-types (\cite{bianchi2017}),
%due the high hydrophobicity as is required for a TMH to span
%the (hydrophobic) cell membrane, combined with the fact that the MHC
%binding cleft is hydrophobic as well.
%
%It is unknown if the presentation of hydrophobic/TMH residues
%is evolutionary selected for.
%One reason to expect a reduced variability (and hence 
%evolutionary conservation) in TMHs, is
%that these are restricted in their evolution
%by the functional requirement to span a lipid bilayer.
%Due to this, the TMHs that pathogens possess, 
%have a lower chance to develop an escape mutation,
%as many mutations will result in a dysfunctional TMH.

However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
One possibility is that the presentation of 
hydrophobic/TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to this, there are constraints in the types of amino acids 
that can be present in the TMH \richel{ref}. 
Therefore, the TMHs of pathogens 
might have a lower chance to develop an escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

% GvdB:
% The text below here is more discussion 
% as it is already interpretation of the findings. 
% I would add one short paragraph, 
% where you reiterate the most important findings:
% In this study, we show that ...(and than basically 
% the abstract in slightly different words and slightly 
% more detail). End with a statement of why it is 
% important (vaccin development based on TMHs?)...

% RJCB: This was suggested for removal, until the Methods:
% \paragraph{Does MHC-I present TMH-derived epitopes from pathogens as often?}
% 
% It is important that MHC-I presents both the peptides of the 
% (healthy) cell, as well as possible pathogen-derived peptides when the
% cell is infected. As described above, MHC-I presents TMH-derived epitopes 
% from the human host more often. It is unknown if MHC-I has the same
% dedication to present epitopes that stem from TMHs derived from
% proteins produced by pathogen, either would the pathogen
% be a virus or 
% a bacterium.
% 
% \paragraph{MHC-II is expected to present TMHs}
% 
% If presentation of TMHs on MHC-I would bring an evolutionary advantage 
% in the recognition of pathogens by the immune system, 
% it would follow that this is equally important for MHC-II, 
% especially as the help of CD4+ T cells is needed for a long lasting CD8+ T cell 
% response (\cite{novy2007cd4}). 
% The mechanism to detect foreign TMHs by MHC-II would be unknown, 
% similar to the discovery that MHC-I presents TMHs.
% 
% \paragraph{Does MHC-II present TMH-derived epitopes from pathogens as often?}
% 
% It is unknown if MHC-II, like MHC-I, 
% presents TMH-derived epitopes as often, either 
% in humans,
% bacteria or viruses.
% 
% \paragraph{Selection undetectable in whole proteome}
% 
% In general, on would hope that evolutionary selection results in
% an immune system that as most attentive for loci that are
% essential for a virus, as these will be most conserved.
% In SARS-CoV-2, for example, there is preliminary evidence that the strongest
% selection pressure is upon residues that changes its 
% virulence (\cite{velazquez2020positive}).
% These loci, however, only account for a small part of a pathogen's proteome.
% Additionally, these essential parts differ widely between pathogens.
% Because of this scarcity and variance in targets, 
% one can imagine that the human immune system 
% is not tailored to detect these sites, 
% as hinted by upon by the aforementioned influenza study.
% 
% \paragraph{Selection may be detectable in TMHs}
% 
% TMHs, on the other hand, also have their function constraints, 
% yet can occur multiple time a pathogen's proteome.
% One can safely assume a pathogen's proteome contains multiple TMHs.
% Therefore, it may be beneficial for the host
% if its immune system would be more attentive towards TMHs.
% And maybe this has already happened: MHC-I already detects hydrophobic
% peptides. This feature, however, may also be caused by selection
% to detect hydrophobic regions in the soluble proteins of pathogens.
% It is unknown, when focusing on TMHs only, if a signal of selection
% can be detected.
% 
% \paragraph{Use of protein data in phylogenetics}
% 
% When using DNA sequences, one can use a skewed rate
% in non- versus synonymous mutation
% to detect the signature of selection (\cite{murrell2015gene}).
% Using AA sequences, however, has its advantages,
% as its is closer to the actual phenotype
% selection acts upon: DNA may never be transcribed to RNA,
% or its RNA may never be translated (\cite{diz2012proteomics}).
% When using a proteome in phylogenetic research,
% we know that the majority of proteins are selected to just
% maintain their function most of the time, where
% is the time spans there is selection, only a few AAs
% can actually increase the 'fitness' of the
% protein (\cite{anisimova2009investigating}).
% There, when generalizing the dynamics of mostly purifying selection (to maintain
% a protein's function) and a short duration of positive selection,
% those genes that are selected cannot be detected (\cite{yang2000statistical}).
